---
title: "figure 4 5'/3'UTR vignettes"
author: "Kent Riemondy RBI"
date: '`r Sys.Date()`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message = FALSE, warning = FALSE, include = FALSE}
source(here::here("R/globals.r"))

fig_dir <- "figs_utrs"
out_dir <- "output_files"
dir.create(fig_dir, recursive = TRUE)
```


```{r}
isos <- read_tsv(file.path(out_dir, "all_isoform_annotations.tsv"))
```

##' Check for Alternative PAS usage

```{r}
isos_split <- filter(isos, !same_three_prime_utr) %>% 
  split(., .$tx_class)

utr_gtf <- tidy_gtf(annots$drosophila$gtf) %>% 
  filter(type == "three_prime_utr",
         transcript_id %in% isos$featureID) 

# filter for isoforms with the same 3'UTR start position, just different ends,
# i.e. don't compare uts generated solely via splicing, rather more likely through APA
comparable_utrs <- utr_gtf %>%
  select(transcript_id, start, end, strand) %>% 
  unique() %>% 
  left_join(select(isos,
                   gene_id,
                   tx_class,
                   featureID),
            by = c("transcript_id" = "featureID")) %>% 
  group_by(transcript_id) %>%
  mutate(tx_start_pos = ifelse(strand == "+",
                           min(start),
                           max(end))) %>% 
  group_by(gene_id) %>% 
  mutate(comp_utr = length(unique(tx_start_pos)) == 1) %>% 
  filter(comp_utr) %>% 
  pull(transcript_id) %>% 
  unique()

#utr_gtf <- filter(utr_gtf, transcript_id %in% comparable_utrs)

seq_df_fwd <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "+") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, start, end, .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df_rev <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "-") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, desc(start), desc(end), .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df <- bind_rows(seq_df_fwd, seq_df_rev)
```

```{r}
canonical_pas <- c(
  "AATAAA",
  "ATTAAA"
)

noncanonical_pas <- c(
  "AGTAAA",
  "TATAAA",
  "TTTAAA",
  "AATATA",
  "AATACA",
  "CATAAA",
  "AAAATA",
  "GATAAA",
  "TATATA",
  "ACTAAA",
  "AGATAA",
  "AATAGA",
  "ATTACA",
  "AGAATA"
)

pas_signals <- lst(canonical_pas,
                 noncanonical_pas)


find_pas_signals <- function(df_w_seqs, pas_signals){
  
  pas_seq_matches <- map_dfr(pas_signals, 
                             function(signals){
                               names(signals) <- signals
                               map_dfr(signals, 
                                       function(x){
                                         res <- find_seq_match(df_w_seqs$seq, x)
                                         names(res) <- df_w_seqs$transcript_id
                                         map_dfr(res, as.data.frame, .id = "transcript_id")
                                       }, .id = "pas")
                             }, .id = "pas_type")
  
  seq_len <- df_w_seqs %>%
    mutate(utr_length = nchar(seq)) %>% 
    select(-seq)
  
  pas_seq_matches <- left_join(pas_seq_matches, seq_len, by = "transcript_id")
  pas_seq_matches <- filter(pas_seq_matches, (utr_length - start) <= 50)
  pas_seq_matches
  
}
```

```{r}
tx_map <- map(isos_split, ~.x$featureID %>% .[. %in% comparable_utrs])

pas_res <- map_dfr(tx_map, 
    ~filter(seq_df, 
           transcript_id %in% .x) %>% 
      find_pas_signals(., pas_signals),
    .id = "tx_class"
    )

plt_all <- ggplot(pas_res, aes(tx_class)) + 
  geom_bar(aes(fill = pas_type), position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "",
       y = "Poly(A) Signal Count",
       fill = "")

plt_can <- filter(pas_res, pas_type == "canonical_pas") %>% 
  ggplot(., aes(tx_class)) + 
  geom_bar(aes(fill = pas), position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "",
       y = "Poly(A) Signal Count",
       fill = "")

plt_noncan <- filter(pas_res, pas_type != "canonical_pas") %>% 
  ggplot(., aes(tx_class)) + 
  geom_bar(aes(fill = pas), position = "fill") +
  scale_fill_manual(values = scbp::discrete_palette_default) +
  labs(x = "",
       y = "Poly(A) Signal Count",
       fill = "")

p <- plot_grid(plt_all, plt_can, plt_noncan,nrow = 1, ncol = 3)

p

save_plot(file.path(fig_dir, "pas_usage.pdf"), 
          p,
          nrow = 1,
          ncol = 3,
          base_asp = 1.2)
```


There appears to be a similar distribution of PAS usage between maternal and zygotic isoforms


## mir-309 motifs

`miR-309` sits in a cluster of miRNAs with 5 distinct seeds https://doi.org/10.1016/j.cub.2008.02.081

mir-3 and mir-309 have the same 2-7mer

```{r}
mir_fa <- c(">dme-miR-309-3p MIMAT0000401",
     "GCACUGGGUAAAGUUUGUCCUA",
    # ">dme-miR-3-3p MIMAT0000108",
  #   "UCACUGGGCAAAGUGUGUCUCA",
     ">dme-miR-6-1-5p MIMAT0020787",
     "AGGGAAUAGUUGCUGUGCUGUA",
     ">dme-miR-286-3p MIMAT0000359",
     "UGACUAGACCGAACACUCGUGCU",
     ">dme-miR-3-3p MIMAT0000108",
     "UCACUGGGCAAAGUGUGUCUCA",
     ">dme-miR-4-3p MIMAT0000109",
     "AUAAAGCUAGACAACCAUUGA",
     ">dme-miR-5-5p MIMAT0000110",
     "AAAGGAACGAUCGUUGUGAUAUG")
mir_seqs <- mir_fa[seq(2, length(mir_fa), 2)]
names(mir_seqs) <- mir_fa[seq(1, length(mir_fa), 2)]

# use bartel lab style seed match conventions
# 8mer = match positions 2-8 with A at position 1
# 7mer-m8 = match positions 2-8 
# 7mer-A1 = match positions 2-7 with A at position 1
# 6mer-A1 = match positions 2-7
mir_seeds <- map(mir_seqs, 
                 ~{
                   seed_match <- str_to_upper(.x) %>% 
                     str_replace_all("U", "T") %>% 
                     str_sub(2, 8) %>% 
                     revComp()
                   mer8 <- str_c(seed_match, "A")
                   mer7m8 <- seed_match
                   mer7a1 <- str_c(str_sub(seed_match, 2), "A")
                  # mer6 <- str_sub(seed_match, 2)
                   lst(mer8,
                       mer7m8,
                       mer7a1)
                 })

find_mir_seeds <- function(df_w_seqs, mir_seeds){
  
  mir_seq_matches <- map_dfr(mir_seeds, 
                             function(seeds){
                               res <- map_dfr(seeds, 
                                       function(x){
                                         res <- find_seq_match(df_w_seqs$seq, x)
                                         names(res) <- df_w_seqs$transcript_id
                                         map_dfr(res, 
                                                 ~{
                                                   out <- as.data.frame(.x)
                                                   if(nrow(out) == 0){
                                                     out <- data.frame(start = 0, 
                                                                       end = 0,
                                                                       match = FALSE)
                                                   } else {
                                                     out$match <- TRUE
                                                   }
                                                   
                                                   out},
                                                   .id = "transcript_id")
                                       }, .id = "seed_type")
                             }, .id = "miRNA")
  
  mir_seq_matches
  
}
```

```{r}
tx_map <- map(isos_split, ~.x$featureID )

mir_res <- map_dfr(tx_map, 
    ~filter(seq_df, 
           transcript_id %in% .x) %>% 
      find_mir_seeds(., mir_seeds),
    .id = "tx_class"
    )
seed_ranking <- c("mer8", "mer7m8", "mer7a1")

mir_res <- group_by(mir_res, tx_class, transcript_id) %>% 
  mutate(any_match_found = any(match),
         seed_type = factor(seed_type, levels = seed_ranking)) %>% 
  arrange(seed_type, .by_group = TRUE) %>% 
  mutate(best_seed = ifelse(any_match_found,
                            as.character(seed_type)[match][1],
                            "no match")) %>% 
  ungroup()

plt_dat <- select(mir_res, tx_class, transcript_id, best_seed) %>% 
  unique()
pretty_seed_labels <- c(
  "mer8" = "8mer",
  "mer7m8" = "7mer-m8",
  "mer7a1" = "7mer-A1",
   "no match" = "No match"
)

plt_dat <- mutate(plt_dat, pretty_label = pretty_seed_labels[best_seed])
plt_prop <- ggplot(plt_dat, aes(tx_class)) + 
  geom_bar(aes(fill = pretty_label), position = "fill") +
  scale_fill_manual(values = scbp::discrete_palette_default) +
  labs(x = "",
       y = "3'UTRs with miR-309\ncluster seed matches",
       fill = "")

plt_prop

save_plot(file.path(fig_dir, "mir309_cluster_seeds.pdf"), 
          plt_prop,
          nrow = 1,
          ncol = 1,
          base_asp = 1.0)

```


## 5'UTR length

5'UTRs tend to increase in length in zygotic isoforms. Is a change in 5'UTR length associated with changes in translational efficiency?

```{r}
te <- read_tsv("../2019-11-21_rpfs/transcript_level_te.tsv.gz")
iso_annots <- read_tsv(file.path(out_dir, "all_isoform_annotations.tsv"))


utr_lengths <- iso_annots %>% 
  filter(same_three_prime_utr, same_CDS, !same_five_prime_utr) %>% 
  select(gene_id, five_prime_utr_length, tx_class) %>% 
  pivot_wider(names_from = tx_class,
              values_from = five_prime_utr_length) %>% 
  mutate(delta_lengths = zygotic - maternal)
  
delta_5utr <- iso_annots %>% 
  filter(same_three_prime_utr, same_CDS, !same_five_prime_utr) %>% 
  pull(featureID)

te_iso_changes <- iso_annots %>%
  filter(featureID %in% delta_5utr) %>% 
  select(tx_class, featureID, gene_name) %>% 
  left_join(te, by = c("featureID" = "transcript_id")) 
  

te_iso_changes <- na.omit(te_iso_changes)
sample_order <- dplyr::select(te, Stage_11:`5-6_h`) %>% 
  colnames()
  
te_changes <- te_iso_changes %>% 
  pivot_longer(cols = -(tx_class:representative_cds_transcript)) %>% 
  mutate(sample = factor(name, levels = sample_order),
         value = log2(value)) %>% 
  select(representative_cds_transcript:sample) %>% 
  unique()


ggplot(te_changes, aes(sample, value)) + 
  geom_violin(aes(fill = sample),
              position = position_dodge()) +
  labs(x = "",
       y = "Translation Efficiency (log2)") +
  scale_fill_brewer(palette = "Set1", 
                    name = "") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )


te_changes <- te %>% 
  pivot_longer(cols = -(transcript_id:representative_cds_transcript)) %>% 
  mutate(sample = factor(name, levels = sample_order),
         value = log2(value)) %>% 
  select(representative_cds_transcript:sample) %>% 
  unique()

te_changes <- na.omit(te_changes)
ggplot(te_changes, aes(sample, value)) + 
  geom_violin(aes(fill = sample),
              position = position_dodge()) +
  labs(x = "",
       y = "Translation Efficiency (log2)") +
  scale_fill_brewer(palette = "Set1", 
                    name = "") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

```

Many 5'UTR changes appear to be driven by splicing rather than changes in 5'UTR start.


```{r}
gtf <- tidy_gtf(annots$drosophila$gtf)

tx_starts <- filter(gtf, type == "transcript") %>% 
  group_by(transcript_id) %>% 
  summarize(tx_start = ifelse(strand == "+",
                              start,
                              end))

utr5_annots <- iso_annots %>% 
  filter(same_three_prime_utr, same_CDS, !same_five_prime_utr) %>% 
  left_join(tx_starts, by = c("featureID" = "transcript_id"))

utr5_annots %>% 
  group_by(gene_id) %>% 
  summarize(same_start_site = ifelse(length(unique(tx_start)) == 1,
                TRUE,
                FALSE))
```



## 3' UTR motifs


```{r}
kmer_enrichment <- function(all_kmers, 
                            group_1_genes,
                            group_2_genes = NULL,
                            id = "transcript_id"){
  
  bg_genes <- unique(all_kmers[[id]])
  test_genes <- intersect(group_1_genes, bg_genes)
  
  if(!is.null(group_2_genes)){
    bg_genes <- intersect(group_2_genes, bg_genes)
  } 

  bg <- all_kmers %>% 
      filter(!!sym(id) %in% bg_genes) %>% 
      mutate(success = ifelse(counts >= 1, 1L, 0L)) %>% 
      group_by(kmer) %>% 
      summarize(counts = sum(success))
  
  kmer_count <- all_kmers %>% 
    filter(!!sym(id) %in% group_1_genes) %>% 
    mutate(success = ifelse(counts >= 1, 1L, 0L)) %>% 
    group_by(kmer) %>% 
    summarize(counts = sum(success))
  
  kmers_total <- left_join(kmer_count, 
                           bg,
                           by = "kmer",
                           suffix = c("_test", "_bg"))
  
  kmers_total <- mutate(kmers_total, 
                        n_seq_test = length(unique(test_genes)),
                        n_seq_bg = length(unique(bg_genes))) %>% 
    na.omit()
  
  out <- kmers_total %>% 
    na.omit(.) %>%
    rowwise() %>% 
    do(broom::tidy(binom.test(.$counts_test, .$n_seq_test, 
                       p = .$counts_bg / .$n_seq_bg,
            alternative = "greater"))) 
  
  res <- bind_cols(kmers_total, out) %>% 
    mutate(padj = p.adjust(p.value),
           prop.test = counts_test / n_seq_test,
           prop.bg = counts_bg / n_seq_bg,
           enrichment = log2(prop.test / prop.bg),
           log10pval = -log10(padj)) %>% 
    arrange(padj) %>% 
    mutate(rank = row_number(),
           enriched = ifelse(rank <= 10, T, F)) %>% 
    arrange(padj)
  
  res
  
}
      


plot_kmers <- function(kmer_summary, 
                       plot_title = "Kmers enriched in promoters",
                       n = "",
                       motif = "CAGGTAG"){
  
  labeled_kmers <- mutate(kmer_summary, 
                          enriched = ifelse(!enriched,
                                            str_detect(kmer, motif) & 
                                              log10pval < -log10(0.05),
                                            enriched)) %>%
                            filter(enriched)
  
  plt <- ggplot(kmer_summary,
                     aes(enrichment, log10pval)) +
  geom_point() +
  geom_text_repel(data = labeled_kmers, 
                  aes(label = kmer), color = "red", 
                  force = 10) +
  labs(title = plot_title, 
       subtitle = paste0(n, " genes"),
       x = "Kmer Enrichment (log2)")
  plt
}

```



```{r}
kmer_size <- 7
isos_split <- filter(isos, !same_three_prime_utr) %>% 
  split(., .$tx_class)

utr_gtf <- tidy_gtf(annots$drosophila$gtf) %>% 
  filter(type == "three_prime_utr")

seq_df_fwd <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "+") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, start, end, .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df_rev <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "-") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, desc(start), desc(end), .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df <- bind_rows(seq_df_fwd, seq_df_rev)

kmers_3utrs <- mutate(seq_df, 
                      kmers = get_kmers(seq,
                                        n = 7,
                                        both_strands = FALSE)) %>% 
                select(transcript_id, kmers) %>% 
  unnest_legacy() 

kmer_results <- kmer_enrichment(kmers_3utrs, 
                      isos_split$zygotic$featureID,
                      isos_split$maternal$featureID)


a <- filter(kmers_3utrs, transcript_id %in% isos_split$zygotic$featureID) %>% 
  pull(kmer)

zyg_kmers <- filter(seq_df, 
                    transcript_id %in% isos_split$zygotic$featureID) %>% 
  pull(seq) %>% 
  unique()

mat_kmers <- filter(seq_df, 
                    transcript_id %in% isos_split$maternal$featureID) %>% 
  pull(seq) %>% 
  unique()
```

```{r}
library(transite)

v_all_dat <- calculateKmerEnrichment(list(zyg_kmers,
                                  mat_kmers),
                             seq_df$seq, k = 6)

v_each_other <- calculateKmerEnrichment(list(zyg_kmers),
                             c(zyg_kmers, mat_kmers), k = 6)

map(v_all_dat$dfs, 
    ~mutate(.x, kmers = v_all_dat$kmers) %>%
      arrange(adj.p.value)) -> a

mutate(v_each_other$dfs[[1]], kmers = v_all_dat$kmers) %>%
      arrange(adj.p.value) -> b


v_each_other <- runKmerTSMA(zyg_kmers,
                             c(zyg_kmers, mat_kmers))
```




## 5' UTR kmers

```{r}
kmer_size <- 7
isos_split <- filter(isos, !same_five_prime_utr) %>% 
  split(., .$tx_class)

utr_gtf <- tidy_gtf(annots$drosophila$gtf) %>% 
  filter(type == "five_prime_utr")

seq_df_fwd <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "+") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, start, end, .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df_rev <- get_sequences(utr_gtf, annots$drosophila$genome) %>% 
  filter(strand == "-") %>% 
    dplyr::group_by(transcript_id) %>% 
    dplyr::arrange(chrom, desc(start), desc(end), .by_group = TRUE) %>% 
    dplyr::summarize(seq = stringr::str_c(seq, collapse = ""))

seq_df <- bind_rows(seq_df_fwd, seq_df_rev)

kmers_5utrs <- mutate(seq_df, 
                      kmers = get_kmers(seq,
                                        n = 7,
                                        both_strands = FALSE)) %>% 
                select(transcript_id, kmers) %>% 
  unnest_legacy() 

isos_split <- split(isos, isos$tx_class)
kmer_results <- kmer_enrichment(kmers_5utrs, 
                      isos_split$zygotic$featureID,
                      isos_split$maternal$featureID)


```

```{r}
library(transite)


zyg_kmers <- filter(seq_df, 
                    transcript_id %in% isos_split$zygotic$featureID) %>% 
  pull(seq) %>% 
  unique()

mat_kmers <- filter(seq_df, 
                    transcript_id %in% isos_split$maternal$featureID) %>% 
  pull(seq) %>% 
  unique()

v_all_dat <- calculateKmerEnrichment(list(zyg_kmers,
                                  mat_kmers),
                             seq_df$seq, k = 6)

v_each_other <- calculateKmerEnrichment(list(zyg_kmers),
                             c(zyg_kmers, mat_kmers), k = 6)

map(v_all_dat$dfs, 
    ~mutate(.x, kmers = v_all_dat$kmers) %>%
      arrange(adj.p.value)) -> a

mutate(v_each_other$dfs[[1]], kmers = v_all_dat$kmers) %>%
      arrange(adj.p.value) -> b


v_each_other <- runKmerTSMA(zyg_kmers,
                             c(zyg_kmers, mat_kmers))
```
