---
title: "Chicken MZT analysis"
author: "Kent Riemondy RBI"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message = FALSE, warning = FALSE, include = FALSE}
source(here::here("R/globals.r"))
fig_dir <- "figs"
species <- "chicken"
dir.create(fig_dir, showWarnings = FALSE)
```


## Approach

A paper describing a very high resolution time course of zebrafish was published in eLife [A high-resolution mRNA expression time course of embryonic development in zebrafish](https://elifesciences.org/articles/30860). This dataset will likely provide additional proof of principle data for the intronic analysis performed with drosophila datasets. 

RNA-Seq data (paired-end) from 8 stages surrounding the MZT (or ZGA in zebrafish literature) was downloaded with 5 replicates per time point. The data was processed through salmon similarly to the drosophila data. In contrast to Olivia's drosphila data, these data are poly(A) purified mRNA-Seq libraries, so the intronic signal may be more subtle.  

 
## get metadata

```{r get_metadata}
mdata <- read_tsv(file.path(data_dir, "raw_data",
                             "chicken", 
                             "HWANG", "PRJNA342320_download_log.txt"))

mdata$study <- 'HWANG'

#cleanup sample names
mdata <- mdata %>% 
  separate(sample_title, c("stage", "replicate"), sep = "_") %>% 
  select(-fastq_aspera, -fastq_ftp)

mdata
```

```{r }
gtf <- tidy_gtf(annots[[species]]$gtf)

gene2tx <- gtf %>%
   filter(type == "exon") %>%
   select(gene_id, transcript_id) %>%
   na.omit() %>%
   unique()

gene2intron <- read_tsv(here("dbases-rerun", species, "eisa", "tx2gene.tsv"))
```

## Load in featurecounts data

```{r load_dat, message = F}
count_data_dir <- file.path(data_dir, 
                            "featurecounts",
                            "chicken")
count_files <- dir(count_data_dir,
                   pattern = "_counts.tsv$",
                   recursive = T,
                   full.names = T)

sample_names <- str_split(count_files, "\\/") %>% 
  map_chr(., ~.x[10]) %>% 
  str_remove("_counts.tsv") 

sample_names_simple <- str_replace(sample_names, "_exon|_intron", "")
 
dat <- map(count_files, 
           ~read_tsv(.x,
                     skip = 2,  
                     col_names = F) %>% 
             select(gene_id = X1, 
                    length = X6, 
                    counts = X7))
             
names(dat) <- sample_names

dat <- bind_rows(dat, .id = "sample_name")

dat <- mutate(
  dat,
  run_accession = str_replace(sample_name,
                    "_exon|_intron",
                    ""),
  count_type = ifelse(str_detect(sample_name, "exon"),
                      "exon",
                      "intron")
)

#join with main data
dat <- inner_join(dat, 
                  mdata,
                  by = c("run_accession"))
```

### transcript to gene mapping

```{r drop_monoexonic}
gtf <- import(annots[[species]]$gtf)
gtf <- as.data.frame(gtf)
gtf$seqnames <- as.character(gtf$seqnames)

gtf_summary <- dplyr::filter(gtf, type == "exon") %>% 
  dplyr::select(seqnames, start, end, strand, gene_id) %>% 
  dplyr::rename(chrom = seqnames) %>% 
  unique() %>% 
  group_by(gene_id) %>% 
  bed_merge() %>% 
  dplyr::summarize(total_exons = n())

monoexonic_genes <- dplyr::filter(gtf_summary, total_exons == 1) %>%
  pull(gene_id)
```


```{r remove_monoexonic}
dat <- anti_join(dat, 
                 tibble(gene_id = monoexonic_genes),
                 by = "gene_id")

dat <- group_by(dat, run_accession) %>% 
  mutate(lib_size = sum(counts),
         eff_length = as.numeric(length) - 150 + 1) %>% 
  filter(eff_length > 0 ) %>% 
  mutate(rate = log(counts) - log(eff_length),
         total_rates = log(sum(exp(rate))),
         tpm = exp(rate - total_rates + log(1e6)))
```

### plot pre_mRNA ratio

```{r, calc_premRNA}
# sum primary and mature TPMs for each gene across replicates
summary_dat <- dplyr::select(dat, tpm, gene_id, 
              run_accession, count_type, stage,
              study, replicate) %>% 
  group_by(gene_id, run_accession, 
           stage, count_type) %>% 
  dplyr::summarize(total_tpm = sum(tpm, na.rm = F)) %>% 
  ungroup() 

## calc primary ratio
tpm_summary <- ungroup(dat) %>% 
  dplyr::select(tpm, gene_id, 
              count_type, stage,
              study, replicate) %>% 
  spread(count_type, tpm, fill = 0L) %>% 
  mutate(primary_ratio  = intron / (intron + exon))


## calc mean primary ratio
tpm_means <- tpm_summary %>% 
  group_by(gene_id, 
           stage) %>% 
  summarize(mean_primary = mean(primary_ratio,
                                na.rm = T)) 

tpm_means <- mutate(tpm_means, 
                       stage = factor(stage, 
                                         levels = c("Oocyte",
                                                    "Zygote",
                                                    "EGKI",
                                                    "EGKIII",
                                                    "EGKVI",
                                                    "EGKVIII",
                                                    "EGKX"
                                                    )
                                         )
                       ) %>% 
  ungroup()

plt <- ggplot(tpm_means, 
       aes(stage, mean_primary)) +
  geom_boxplot(aes(fill = stage), coef = 1e3) + 
  labs(x = "",
       y = expression(frac("pre-mRNA TPM", 
                           "mature + pre-mRNA TPM"))) +
  scale_fill_brewer(palette = "Set1",
                    name = "Transcript Type",
                    breaks = c("pre", "mature"),
                    labels = c("pre-mRNA", "mature-mRNA")) +
  theme(
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5,
                               hjust = 0.5)
  )
# 
# save_plot("chicken-pre-mRNA-ratios.pdf", plt, 
#           base_width = 6,
#           base_height = 4.5)
plt

```

## Examine pre-mRNA ratios

```{r read_in_salmon}
fa_types <- c(
  "eisa_masked"
)

files <- map(fa_types, 
             ~file.path(data_dir,
                        "salmon_bt2_masked", 
                       species, 
                      "HWANG",
                        .x,
                        mdata$run_accession,
                        "quant.sf"))

dat <- map(files, ~map(.x, read_tsv))
names(dat) <- fa_types

dat <- map(dat, function(x) {
  names(x) <- mdata$run_accession
  x
} )

dat <- map(dat, ~bind_rows(.x, .id = "run_accession"))

dat <- bind_rows(dat, .id = "expt")

#join with main data
dat <- inner_join(dat, 
                 mdata,
                by = c("run_accession"))

# classify transcripts as primary or mature
dat <- dat %>% 
  mutate(count_type = ifelse(str_detect(Name, 
                                        "-I[0-9]*$"), 
                             "intron", 
                             "exon"))

# drop unneeded data
dat <- dplyr::select(dat,
                     -c(Length:EffectiveLength, 
                        NumReads))

dat <- mutate(dat,
              tmp_id = ifelse(count_type == "intron",
                               str_remove(Name, "-I[0-9]*$"),
                               Name))
# add in geneid attribute
tpm_dat <- left_join(dat, gene2tx, 
                 by = c("tmp_id" = "transcript_id")) %>% 
  dplyr::rename(transcript_id = Name)

```

```{r}
# sum primary and mature TPMs for each gene 
summary_dat <- dplyr::select(tpm_dat, TPM, gene_id, 
              run_accession, count_type, replicate, stage) %>% 
  group_by(gene_id, run_accession, replicate,
           stage, count_type) %>% 
  dplyr::summarize(total_tpm = sum(TPM, na.rm = F)) %>% 
  ungroup() 

## calc primary ratio
tpm_summary <- summary_dat %>% 
  spread(count_type, total_tpm) %>% 
  mutate(primary_ratio  = intron / (intron + exon))

## calc mean primary ratio
tpm_means <- tpm_summary %>% 
  group_by(gene_id, stage) %>% 
  summarize(mean_primary = mean(primary_ratio, na.rm = T)) 

tpm_means <- mutate(tpm_means, 
                       stage = factor(stage, 
                                         levels = c("Oocyte",
                                                    "Zygote",
                                                    "EGKI",
                                                    "EGKIII",
                                                    "EGKVI",
                                                    "EGKVIII",
                                                    "EGKX"
                                                    )
                                         )
                       ) %>% 
  ungroup()

tpm_means <- dplyr::ungroup(tpm_means)

p <- ggplot(tpm_means, 
       aes(stage, mean_primary)) +
  geom_boxplot(aes(fill = stage), outlier.shape = NA, coef = 1e10) + 
  labs(x = "",
       y = expression(frac("pre-mRNA TPM", 
                           "mature + pre-mRNA TPM"))) +
  scale_fill_brewer(palette = "Paired") +
  geom_vline(xintercept = 1.5, linetype = 'dashed') + 
  geom_vline(xintercept = 4.5, linetype = 'dashed') + 
  theme(
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5,
                               hjust = 0.5),
    legend.position = "none"
  )

save_plot(file.path(fig_dir, "chicken-pre-mRNA-ratios.pdf"),
                    p,
          base_width = 6,
          base_height = 4.5)

p

save_plot( "../figures/pdfs/fig1/chicken-pre-mRNA-ratios.pdf",
           p,
          base_width = 6,
          base_height = 4.5)

p
```


```{r}
sessioninfo::session_info()
```
