---
title: "Per transcript analysis"
author: "Kent Riemondy RBI"
date: '`r Sys.Date()`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message = FALSE, warning = FALSE, include = FALSE}
source(here::here("R/globals.r"))

fig_dir <- "figs_eisa"
dir.create(fig_dir, recursive = TRUE)
```


## Approach

Each total RNA-Seq lib was trimmed to remove Tru-Seq adapters (Rissland libraries), or barcodes and custom 3' adapter (Bartel libraries). The trimmed reads were then processed with `salmon` to estimate `TPMs` for each transcript. The `salmon` transcript database was modified to include a pre-mRNA entry, which was designated as an exon that spanned the entire transcript This pre-mRNA annotation matches the `transcript` entry from the annotation file. By including the pre-mRNA transcript in the `salmon` database, one can get an estimate of the fraction of pre-mRNA or mature mRNA present in each sample. 

To make this approach more accurate it is necessary to mask intronic signals that are present in the transcriptionally quiescent pre-MZT stages (oocytes or 0 hours). This is done by first aligning the data to the genome, using STAR, then extracting out intronic regions with coverage in the 0 hour timepoint. These regions are then masked (converted to Ns) to exclude counting these regions when quantifying intronic reads. 

Salmon inserts psuedorandom nucleotides into N's during index generation, which is potentially problematic. Instead the reads are aligned with bowtie2 with a high -k setting to report many multiple alignments. Bowtie will treat N's as missing sequence. Following alignment, salmon is used for quantification. 

## get metadata

```{r get_metadata}

mdata1 <- read_tsv(file.path(data_dir,
                             "raw_data",
                             "drosophila",
                             "EICHHORN",
                             "GSE83616_run_info_geo.txt"))

mdata2 <- read_tsv(file.path(data_dir,
                             "raw_data",
                             "drosophila",
                             "RISSLAND",
                             "GSE98106_run_info_geo.txt"))

#simplify mdata
mdata1 <- dplyr::select(mdata1,
                       Run_s, 
                       developmental_stage_s,
                       genotype_s,
                       source_name_s,
                       Assay_Type_s)

#simplify mdata
mdata2 <- dplyr::select(mdata2,
                       Run_s, 
                       developmental_stage_s,
                       rip_antibody_s,
                       source_name_s,
                       strain_s)

# select relevant libs
mdata2 <- dplyr::filter(mdata2,
                     source_name_s == "Embryo",
                     strain_s %in% c("w1118", "eGFP-ME31B"),
                     rip_antibody_s == "none") %>% 
  dplyr::select(-rip_antibody_s)

mdata1 <- dplyr::filter(mdata1,
                     genotype_s == "wt",
                     Assay_Type_s == "RNA-Seq") %>% 
  dplyr::select(-Assay_Type_s) %>% 
  dplyr::rename(strain_s = genotype_s)


#mdata <- list(mdata1, mdata2)
#names(mdata) <- c("EICHHORN", "RISSLAND")
mdata <- list(mdata2)
names(mdata) <- c("RISSLAND")

mdata <- bind_rows(mdata, .id = "study")
#cleanup names
mdata <- mdata %>% 
  mutate(source_name_s = ifelse(source_name_s == "embryos", 
                                "Embryo",
                                source_name_s),
         developmental_stage_s = str_replace(developmental_stage_s,
                                             " h ",
                                             " hr "))

# cleanup colnames
colnames(mdata) <- str_replace(colnames(mdata),
                               "_s$",
                               "")

mdata

#dir.create("processed_files")
#write_tsv(mdata, "processed_files/metadata.tsv")
```

```{r }
gtf_fn <- "~/Projects/shared_dbases/annotation/drosophila/Drosophila_melanogaster.BDGP6.84.gtf"

gtf <- tidy_gtf(gtf_fn)

gene2symbol <- gtf %>%
  filter(type == "gene") %>% 
  select(gene_id, gene_name) %>% 
  unique()

gene2tx <- gtf %>%
   filter(type == "transcript") %>%
   select(gene_id, transcript_id) %>%
   na.omit() %>%
   unique()

gene2intron <- read_tsv("annotations/eisa_tx2gene.tsv")
```


## Examine pre-mRNA ratios

```{r read_in_salmon}

fa_types <- c(
  "eisa_masked"
)


files <- map(fa_types, 
             ~file.path(data_dir,
                        "salmon", 
                        "drosophila",
                        mdata$study,
                        .x,
                        mdata$Run,
                        "quant.sf"))

dat <- map(files, ~map(.x, read_tsv))
names(dat) <- fa_types

dat <- map(dat, function(x) {
  names(x) <- mdata$Run
  x
} )

dat <- map(dat, ~bind_rows(.x, .id = "Run"))

dat <- bind_rows(dat, .id = "expt")

#join with main data
dat <- inner_join(dat, 
                 mdata,
                by = c("Run"))

# classify transcripts as primary or mature
dat <- dat %>% 
  mutate(count_type = ifelse(str_detect(Name, 
                                        "-I[0-9]*$"), 
                             "intron", 
                             "exon"))

# drop unneeded data
dat <- dplyr::select(dat,
                     -c(Length:EffectiveLength, 
                        NumReads))

dat <- mutate(dat,
              tmp_id = ifelse(count_type == "intron",
                               str_remove(Name, "-I[0-9]*$"),
                               Name))
# add in geneid attribute
tpm_dat <- left_join(dat, gene2tx, 
                 by = c("tmp_id" = "transcript_id")) %>% 
  dplyr::rename(transcript_id = Name)

```

```{r}
# sum primary and mature TPMs for each gene 
summary_dat <- dplyr::select(tpm_dat, TPM, gene_id, 
              Run, count_type, developmental_stage,
              study, strain) %>% 
  group_by(gene_id, Run, 
           developmental_stage, count_type,
           strain) %>% 
  dplyr::summarize(total_tpm = sum(TPM, na.rm = F)) %>% 
  ungroup() 

## calc primary ratio
tpm_summary <- summary_dat %>% 
  spread(count_type, total_tpm) %>% 
  mutate(primary_ratio  = intron / (intron + exon))


## calc mean primary ratio
tpm_means <- tpm_summary %>% 
  mutate(
    lib_type = ifelse(
      strain == "wt",
      "Poly(A)+",
      "Ribo-Zero"
    )
  ) %>% 
  group_by(gene_id, 
           developmental_stage,
           lib_type) %>% 
  summarize(mean_primary = mean(primary_ratio,
                                na.rm = T)) 

dev_stages_ordered <- c(
  "Stage 11 oocyte",
  "Stage 12 oocyte",
  "Stage 13 oocyte",
  "Stage 14 oocyte",
  "Activated egg",
  "0-1 hr embryo",
  "1-2 hr embryo",
  "2-3 hr embryo",
  "3-4 hr embryo",
  "4-5 hr embryo",
  "5-6 hr embryo")

tpm_means <- mutate(tpm_means, 
                       stage = factor(developmental_stage, 
                                      levels = dev_stages_ordered)
                       )


p <- ggplot(tpm_means, 
       aes(stage, mean_primary)) +
  geom_boxplot(aes(fill = stage), outlier.shape = NA, coef = 1e10) + 
  facet_grid(~lib_type, scale = "free") +
  labs(x = "",
       y = expression(frac("pre-mRNA TPM", 
                           "mature + pre-mRNA TPM"))) +
  scale_fill_brewer(palette = "Paired") +
  theme(
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5,
                               hjust = 0.5),
    legend.position = "none"
  )

save_plot(file.path(fig_dir, "eisa-fly-pre-mRNA-ratios.pdf"),
                    p, 
          base_width = 6,
          base_height = 4.5)

p
```

## DE testing

First read in salmon data into tximport.

```{r}
pdata <- mutate(mdata, 
                mzt = ifelse(str_detect(developmental_stage,
                                        "^[2345]-[3456] hr embryo"),
                             "zygotic",
                             "maternal"),
                strain_id = str_replace(strain, "-", "_"))

pdata_split <- split(pdata, pdata$study)

# eichorn_files <- map(fa_types, 
#                  ~file.path(data_dir,
#                    "salmon", 
#                    "drosophila",
#                    pdata_split$EICHHORN$study,
#                    .x,
#                    pdata_split$EICHHORN$Run,
#                    "quant.sf"))

rissland_files <- map(fa_types, 
                 ~file.path(data_dir,
                   "salmon", 
                   "drosophila",
                   pdata_split$RISSLAND$study,
                   .x,
                   pdata_split$RISSLAND$Run,
                   "quant.sf"))

tids <- read_tsv(rissland_files[[1]][1]) %>% 
  dplyr::select(transcript_id = Name) %>% 
  mutate(tmp_id = ifelse(str_detect(transcript_id, "-I[0-9]*$"),
                         str_remove(transcript_id, "-I[0-9]*$"),
                        transcript_id))
                                 
rs_tx_objs <- map(rissland_files, 
               ~tximport(.x, 
                         type = "salmon",
                         tx2gene = gene2intron))

#                                  
# ei_tx_objs <- map(eichorn_files, 
#                ~tximport(.x, 
#                          type = "salmon",
#                          tx2gene = gene2intron))

```

```{r}
# write out tpm matrix
cols <- pdata %>% 
  filter(study == "RISSLAND") %>% 
  mutate(cnames = str_c(developmental_stage, strain, sep = "_")) %>% 
  pull(cnames)
tpm_mat <- rs_tx_objs[[1]]$abundance
colnames(tpm_mat) <- cols

tpm_matrix_rissland <- tidy_matrix(tpm_mat, "gene_id")

outdir <- "processed_files"
dir.create(outdir)
write_tsv(tpm_matrix_rissland, 
          file.path(outdir, "rissland_eisa_tpm_matrix.tsv.gz"))

# write out mean tpm matrix
mean_tpms_rissland <- tpm_matrix_rissland %>% 
  gather(sample, tpm, -gene_id) %>% 
  separate(sample, c("stage", "strain"), sep = "_") %>% 
  group_by(gene_id, stage) %>% 
  summarize(mean_tpm = mean(tpm)) %>% 
  spread(stage, mean_tpm)

# check for annotation issue with duplicate transcripts dropped

genes_to_exclude <- mean_tpms_rissland %>% 
  ungroup() %>% 
  mutate(tx_class = ifelse(str_detect(gene_id, "-I[0-9]*$"),
                             "intron",
                             "exon"),
           gene_id = str_remove(gene_id, "-I[0-9]*$")) %>% 
  pivot_longer(cols = -c(gene_id, tx_class)) %>% 
  pivot_wider(names_from = "tx_class", values_from = "value") %>% 
  filter(is.na(exon)) %>% 
  pull(gene_id) %>% 
  unique()
mean_tpms_rissland <- filter(mean_tpms_rissland, 
                             !gene_id %in% genes_to_exclude,
                             !gene_id %in% str_c(genes_to_exclude, "-I[0-9]*$"))
write_tsv(mean_tpms_rissland, 
          file.path(outdir, "rissland_eisa_tpm_means.tsv.gz"))
```

Run deseq with post-mzt v. pre-mzt
```{r}
to_keep <- mean_tpms_rissland$gene_id

rs_tx_objs <- map(rs_tx_objs, function(x){
  x$abundance <- x$abundance[rownames(x$abundance) %in% to_keep, ]
  x$counts <- x$counts[rownames(x$counts) %in% to_keep, ]
  x$length <- x$length[rownames(x$length) %in% to_keep, ]
  x$infReps <- map(x$infReps, ~.x[rownames(.x) %in% to_keep, ])
  x
})

rs_ddobjs <- map(rs_tx_objs, 
              function(x) {
                dds <- DESeqDataSetFromTximport(x, 
                                                pdata_split$RISSLAND, 
                                                ~strain_id + mzt)
                dds <- DESeq(dds)
                dds
              })

names(rs_ddobjs) <- fa_types
```

Extract results and make MA plots

```{r}
ris_res <- map_dfr(rs_ddobjs, ~results(.x, 
                                       name = "mzt_zygotic_vs_maternal", 
                                       tidy = T), 
                   .id = "fasta_type") %>% 
  tbl_df() %>% 
  mutate(count_type = ifelse(str_detect(row, "-I[0-9]*$"), 
                        "intron",
                        "exon"),
         sig = ifelse(padj < 0.05 & log2FoldChange > 0, "Up", 
                      ifelse(padj < 0.05 & log2FoldChange < 0,
                             "Down", "Not Sig"))) %>% 
  filter(!is.na(sig))

out_tbl <- ris_res %>% 
   mutate(gene_id = str_remove(row, "-I[0-9]*$")) %>% 
  left_join(gene2symbol, by = "gene_id")

write_tsv(out_tbl, file.path(outdir, "rissland_eisa_deseq_results.tsv.gz"))
```

```{r}
plot_ma <- function(df, plt_title = NULL) {
  n_sig_up <- nrow(dplyr::filter(df, sig == "Up")) %>%
    formatC(big.mark = ",")
  n_sig_down <-
    nrow(dplyr::filter(df, sig == "Down"))  %>%
    formatC(big.mark = ",")
  n_notsig <-
    nrow(dplyr::filter(df, sig == "Not Sig"))  %>%
    formatC(big.mark = ",")
  df <- mutate(df, 
               sig = factor(sig, 
                               levels = c("Down",
                                          "Not Sig",
                                          "Up")))
  pre_plot <- ggplot(df,
                     aes(baseMean,
                         log2FoldChange)) +
    geom_point(aes(colour = sig), size = 0.1) +
    scale_colour_viridis(
      discrete = TRUE,
      name = "",
      drop = FALSE,
      labels = c(
        paste0("Significantly decreased (n = ", n_sig_down, ")"),
        paste0("Not Significant (n= ", n_notsig, ")"),
        paste0("Significantly increased (n = ", n_sig_up, ")")
      )
    ) +
    facet_wrap(~count_type) +
    scale_x_log10() +
    labs(x = expression(paste("log"[10], " Average Expression")),
         y = expression(paste("log"[2], " ", frac(
           "Zygotic", "Maternal"
         ))),
         title = plt_title) +
    guides(colour = guide_legend(nrow = 3,
                                 override.aes = list(size = 3))) +
    theme(legend.position = "top")
  
  pre_plot
}


plts <- imap(split(ris_res, 
           list(ris_res$fasta_type,
                         ris_res$count_type)), 
    ~plot_ma(.x, .y))

plt <- plot_grid(plotlist = plts)
save_plot(file.path(fig_dir, 
                    "ma_eisa_plot_salmon.pdf"), plt, nrow = 2, ncol = 2)

plt

```

Make prettier plot

```{r}
ris_res <- filter(ris_res, fasta_type == fa_types)

mean_tpms <- read_tsv("processed_files/rissland_eisa_tpm_means.tsv.gz")

mean_intronic_tpms <- mean_tpms %>% 
  filter(str_detect(gene_id, "-I[0-9]*$")) %>% 
  left_join(ris_res, by = c("gene_id" = "row"))

mean_exonic_tpms <- mean_tpms %>% 
  filter(!str_detect(gene_id, "-I[0-9]*$")) %>% 
  left_join(ris_res, by = c("gene_id" = "row"))

tpm_cut_off <- 0.5
intron_tpm_cut_off <- 0.1
mean_exonic_tpms <- mean_exonic_tpms %>% 
  select(gene_id:fasta_type, -fasta_type) %>% 
  gather(stage, tpm, -gene_id) %>% 
  mutate(expressed = tpm > tpm_cut_off) %>% 
  group_by(gene_id) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  summarize(expr_tpt = stage[expressed][1],# get first tpt above tpm threshold
            max_tpm  = max(tpm, na.rm = TRUE),
            mean_tpm = mean(tpm, na.rm = TRUE)) %>% 
  left_join(mean_exonic_tpms, ., by = "gene_id")

mean_intronic_tpms <- mean_intronic_tpms %>% 
  select(gene_id:fasta_type, -fasta_type) %>% 
  gather(stage, tpm, -gene_id) %>% 
  mutate(expressed = tpm > intron_tpm_cut_off) %>% 
  group_by(gene_id) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  summarize(expr_tpt = stage[expressed][1], # get first tpt above tpm threshold
            max_tpm  = max(tpm, na.rm = TRUE),
            mean_tpm = mean(tpm, na.rm = TRUE)) %>% 
  left_join(mean_intronic_tpms, ., by = "gene_id")

tmp_plt_dat <- bind_rows(mean_intronic_tpms, mean_exonic_tpms) %>% 
  mutate(count_type = factor(count_type, 
                             levels = c("exon", "intron"),
                             labels = c("mRNA", "pre-mRNA"))) %>% 
  arrange(count_type)

ma_plot <- ggplot(tmp_plt_dat %>% filter(padj < 0.05),
                    aes(mean_tpm,
                        log2FoldChange)) +
    geom_point(aes(colour = count_type), size = 0.1) +
    scale_colour_viridis(
      discrete = TRUE,
      name = "",
      drop = FALSE,
    ) +
    scale_x_log10( labels = scales::comma) +
    labs(x = " Mean TPM",
         y = expression(paste("log"[2], " ", frac(
           "Zygotic", "Maternal"
         )))) +
    guides(colour = guide_legend(nrow = 1,
                                 override.aes = list(size = 3))) +
    theme(legend.position = "top")

ma_plot

save_plot(file.path(fig_dir, "ma_plot_eisa_count_type.pdf"), 
          ma_plot, 
          base_aspect_ratio = 1,
          base_height = 5)
```


## Select catagories

```{r}
fold_change_cutoff <- 1
padj_cutoff <- 0.01

maternal_genes <- mean_exonic_tpms %>% 
  filter(`0-1 hr embryo` >= tpm_cut_off) %>% 
  pull(gene_id) %>% 
  unique()

maternal_genes <- mean_intronic_tpms %>% 
  filter(`0-1 hr embryo` >= intron_tpm_cut_off) %>% 
  pull(gene_id) %>% 
  str_remove("-I[0-9]*$") %>% 
  unique() %>% 
  c(., maternal_genes) %>% 
  unique()

zygotic_genes_exon <- mean_exonic_tpms %>% 
     filter(padj < padj_cutoff, 
            log2FoldChange > fold_change_cutoff,
            max_tpm > tpm_cut_off) %>% 
  arrange(padj) %>% 
  pull(gene_id) %>% 
  unique()

zygotic_genes_intron <- mean_intronic_tpms %>% 
     filter(padj < 0.05, 
            log2FoldChange > 0,
            max_tpm > intron_tpm_cut_off) %>% 
    arrange(padj) %>%
  pull(gene_id) %>% 
  unique()

all_zygotic_combined <- unique(union(zygotic_genes_exon,
                                     str_remove(zygotic_genes_intron, "-I[0-9]*$") ))
                              
pure_zygotic_combined <- setdiff(all_zygotic_combined, maternal_genes)

mat_zyg_combined<- intersect(all_zygotic_combined, maternal_genes)

pure_maternal <- setdiff(setdiff(maternal_genes, 
                                 mat_zyg_combined),
                         all_zygotic_combined)

our_gene_lists <- lst(
  all_zygotic_combined,
  pure_zygotic_combined,
  zygotic_genes_intron = str_remove(zygotic_genes_intron, "-I[0-9]*$"),
  zygotic_genes_exon,
  mat_zyg_combined,
  pure_maternal
)

gene_class_df <- our_gene_lists %>% 
  map_dfr(~tibble(gene_id = .x), .id = "mzt_class") %>% 
  left_join(gene2symbol)
```


## Plot out expression

```{r}
gene_class_selected <- filter(gene_class_df, 
                              mzt_class %in% c(
                                "pure_maternal",
                                "pure_zygotic_combined",
                                "mat_zyg_combined"
                              ))

rs_exon_intron_comp <- tpm_dat %>% 
  filter(study == "RISSLAND") %>% 
  group_by(gene_id, developmental_stage, count_type) %>% #average replicate timepoints
  summarize(TPM = mean(TPM)) %>% 
  ungroup() %>% 
  spread(count_type, TPM, fill = 0L)

rs_exon_intron_comps <- list()
rs_exon_intron_comps$exon <- rs_exon_intron_comp %>% 
  select(-intron) %>% 
  spread(developmental_stage, exon) %>% 
  as.data.frame() %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix()

rs_exon_intron_comps$intron  <- rs_exon_intron_comp %>% 
    select(-exon) %>% 
  spread(developmental_stage, intron) %>% 
  as.data.frame() %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix()

rs_exon_intron_comps_log <- map(rs_exon_intron_comps, 
                            ~log1p(.x))

rissland_mat_annot <- rs_exon_intron_comp %>% 
  left_join(gene_class_selected, by = "gene_id") %>% 
  ungroup() %>% 
  select(gene_id, mzt_class) %>% 
  unique() %>% 
  left_join(data_frame(gene_id = rownames(rs_exon_intron_comps_log$exon)), # reorder to match matrices above
            ., by = "gene_id") %>% 
  filter(!is.na(mzt_class))

# only keep genes annotated by intron class
rissland_mat_annot <- rissland_mat_annot %>% 
  filter(!is.na(mzt_class))


rs_exon_intron_comps_log <- map(rs_exon_intron_comps_log, 
                          ~.x[rownames(.x) %in% rissland_mat_annot$gene_id, ])

pretty_labels <- c(
  "pure_maternal" = "Maternal",
  "mat_zyg_combined" = "Maternal + Zygotic",
  "pure_zygotic_combined" = "Zygotic"
)

rissland_mat_annot <- mutate(rissland_mat_annot, 
                             mzt_class = factor(mzt_class, 
                                                levels = names(pretty_labels), 
                                                labels = pretty_labels))

## All classes
h1 <- Heatmap(rs_exon_intron_comps_log$exon, 
                            na_col = "black",
              name = "Exon Abundance",
        show_row_dend = F, 
        show_column_dend = F,
        cluster_columns = FALSE,
        show_row_names = FALSE,
                column_title = "Exon Expression",
        split = rissland_mat_annot$mzt_class)

h2 <- Heatmap(rs_exon_intron_comps_log$intron, name = "Intron Abundance",
              na_col = "black",
        show_row_dend = F, 
        show_column_dend = F,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        column_title = "Intron Expression",
        split = rissland_mat_annot$mzt_class)

pdf(file.path(fig_dir, "eisa_all_classes.pdf"))
draw(h1 + h2)
dev.off()

draw(h1 + h2)
```

## per gene plots

some ideas from de renzis et al.

snail Zygotic Absent Up (;170-fold) Zygotic (;176-fold)
kruppel Zygotic Absent Up (;442-fold) Zygotic (;568-fold)
runt Zygotic Absent Up (;48-fold) Zygotic (;27-fold)
arm Maternal-zygotic Present Up (;2-fold) Zygotic (;2-fold)
fog Maternal-zygotic Present Down (;2-fold) Zygotic (;14-fold)
hb Maternal-zygotic Present Down (;4-fold) Zygotic (;3-fold)
hsp83 Maternal-zygotic Present Down (;16-fold) Zygotic (;8-fold)
nanos Maternal Present Down (;34-fold) Not zygotic
stringd Maternal Present Down (;10-fold) Not zygotic
twine Maternal Present Down (;8-fold) Not zygotic


```{r}

rs_exon_intron_comps_symbol <- map(rs_exon_intron_comps, 
    ~tidy_matrix(.x, "gene_id") %>% 
      left_join(gene2symbol, by = "gene_id") %>% 
      dplyr::select(gene_name, everything(), -gene_id) %>% 
      as.data.frame() %>% 
      column_to_rownames("gene_name") %>% 
      as.matrix())

to_plot <- c(
  "Egfr",
  "Dll",
  "msl-2",
  "smg",
  "Pmi",
  "bic")
cols <- viridis::viridis(3)
plts <- map2(to_plot, rep(cols, each = 2), 
    ~plot_exon_intron_expr(.x, rs_exon_intron_comps_symbol, line_width = 1.5, line_col = .y))


plt <- plot_grid(plotlist = plts, nrow = 3,
          ncol = 2)

plt

save_plot(file.path(fig_dir, "exon_intron_example_genes.pdf"),
          plt,
          nrow = 3, ncol = 2, base_asp = 0.85,
          base_height = 3.5)
```



## Compare to bartel lab data


```{r, eval = F}
# S1 table
url <- "https://genome.cshlp.org/content/suppl/2019/06/21/gr.242164.118.DC1/Supplemental_Table_S1_.xlsx"
fn <- "../../docs/bartel_2019_s1.xlsx"

if(!file.exists(fn)){
  download.file(url, fn)
} 

stages <- c(
  "NC_7-9",
  "NC_9-10",
  "SYNCYTIAL_BLASTODERM",
  "CELLULARIZED_BLASTODERM")

b_dat <- read_excel(fn, skip = 2)

b_genes <- b_dat %>%
  dplyr::select(GENES) %>% 
  left_join(gene2symbol, by = c("GENES" = "gene_name")) %>% 
  na.omit() 

early_zyg <- read_delim("../../docs/derenzis_sup_tbl_8.txt", col_names = F, skip = 1, delim = " ")
early_zyg <- dplyr::select(early_zyg, gene_id = X2) %>% 
  left_join(gene2symbol, by = "gene_id")

all_zyg <- read_delim("../../docs/derenzis_sup_tbl_1.txt", col_names = F, skip = 2, delim = " ")
all_zyg <- dplyr::select(all_zyg, gene_id = X2, og_gene_name = X3) %>% 
  left_join(gene2symbol, by = "gene_id")

mat_zyg <- read_delim("../../docs/derenzis_sup_tbl_6.txt", col_names = F, skip = 2, delim = " ")
mat_zyg <- dplyr::select(mat_zyg, gene_id = X2) %>% 
  left_join(gene2symbol, by = "gene_id")

to_comp <- list(
  `De Renzis et al` = all_zyg$gene_id,
  `Kwasnieski et al` = b_genes$gene_id,
  `Intron method` = gene_class_df %>% filter(mzt_class == "all_zygotic_combined") %>% pull(gene_id)
) %>% 
  map(~na.omit(.x) %>% .[!duplicated(.)])

library(eulerr)

a <- euler(to_comp)

p <- plot(a, 
          quantities = list(fontsize = 16), 
          labels = list(fontsize = 20))
p
pdf(file.path(fig_dir, "classification_overlap.pdf"))
p
dev.off()

all_vals <- unique(unlist(to_comp))

intersection_summary <- imap_dfc(to_comp, 
     ~tibble(!!sym(.y) := all_vals %in% .x )) %>% 
  mutate(gene_id = all_vals) %>% 
  left_join(gene2symbol, by = "gene_id") %>% 
  left_join(out_tbl) 

# 
# intersection_summary %>% 
#   filter(`Intron method`,
#          !`De Renzis et al`,
#          !`Kwasnieski et al`) %>% 
#   filter(padj < 0.00001) %>% 
#   group_by(gene_name) %>% 
#   mutate(n = n()) %>% 
#  filter(n == 1, count_type == "intron") %>% 
#   arrange(padj) %>% 
#   View()
```


## Compare to zelda chip


```{r eisen_dat, eval = F}
if(!file.exists("../../docs/eisen_zelda_chip.tsv")){
  download.file("https://doi.org/10.1371/journal.pgen.1002266.s007", "../../docs/eisen_zelda_chip.tsv")
  download.file("https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3197655/bin/pgen.1002266.s008.xls", 
                "../../docs/eisen_zelda_chip_master_table.xlsx")
}

zelda_chip <- read_tsv("../../docs/eisen_zelda_chip.tsv", 
              col_names = T, 
              skip = 1)

zelda_peaks <- dplyr::select(zelda_chip,
                             chrom = Chr, 
                             start = From,
                             end = To,
                             gene = TSS_gene,
                             dist = TSS_dist,
                             score = Height)

zelda_10k <- zelda_peaks %>% 
  filter(dist <= 10000) %>% 
  left_join(gene2symbol, by = c("gene" = "gene_name"))
  
  
zelda_10k_summed <- group_by(zelda_10k, gene, gene_id) %>%
  summarize(binding_sites = n(),  
            score = max(score)) %>% 
  ungroup()

# use combined data

zelda_chip <- read_xls("../../docs/eisen_zelda_chip_master_table.xlsx") %>% 
  dplyr::select(Gene, `ZLD Max cyc8`:`ZLD Max cyc14`) %>% 
  pivot_longer(cols = -Gene) %>% 
  group_by(Gene) %>% 
  summarize(zelda_signal = max(value)) %>% 
  left_join(gene2symbol, by = c("Gene" = "gene_name"))

plt_dat <- left_join(gene_class_selected, zelda_chip, by = "gene_id") %>% 
  mutate(mzt_class = factor(mzt_class, levels = names(pretty_labels),
                            labels = pretty_labels))
  
p <- ggplot(plt_dat, aes(mzt_class, zelda_signal)) + 
  geom_boxplot(aes(fill = mzt_class),
               coef = Inf) +
  scale_y_log10() +
  labs(x = "",
       y = "Zelda Chip-Seq Signal") +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

save_plot(file.path(fig_dir, "zelda-binding.pdf"),
          p,
          base_asp = 1.2)
```


```{r}
write_tsv(gene_class_df, "processed_files/eisa_gene_classes.tsv")
```



```{r}
outdir <- "processed_files"
gene_class_df <- read_tsv(file.path(outdir, "eisa_gene_classes.tsv"))

mean_tpms_rissland <- read_tsv(file.path(outdir, "rissland_eisa_tpm_means.tsv.gz"))

```




## Annotate time of first expression 

### Based on exonic read increases 

```{r order_by_time_expr_exonic}
gene_class_selected <- filter(gene_class_df, 
                              mzt_class %in% c(
                                 "zygotic_genes_exon"
                              ))
sig_pre_gene_counts <- mean_tpms_rissland %>% 
  filter(str_detect(gene_id, "-I[0-9]*$"),
         str_remove(gene_id, "-I[0-9]*$") %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = str_remove(gene_id, "-I[0-9]*$")) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix()

sig_pre_gene_log_counts <- sig_pre_gene_counts %>% 
  log1p()

sig_gene_log_counts <- mean_tpms_rissland %>% 
  filter(!str_detect(gene_id, "-I[0-9]*$"),
         gene_id %in% gene_class_selected$gene_id) %>% 
 # mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() %>% 
  log1p()


time_to_expr <- sig_pre_gene_counts %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>% 
  gather(stage, expr, -gene) %>% 
  as_tibble() %>% 
  mutate(expressed = expr > 0.1) %>% 
  mutate(stage = factor(stage,
                        levels = c(
                                           "0-1 hr embryo",
                                           "1-2 hr embryo",
                                           "2-3 hr embryo",
                                           "3-4 hr embryo",
                                           "4-5 hr embryo")
                                         )) %>% 
  group_by(gene) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  ungroup() %>% 
  filter(expressed) %>% 
  group_by(gene) %>% 
  summarize(stage_expression = dplyr::first(stage))
            
time_to_expr <- left_join(data_frame(gene = rownames(sig_pre_gene_log_counts)),
                          time_to_expr, by = "gene") %>% 
  na.omit()

ht_pre <- Heatmap(sig_pre_gene_log_counts[time_to_expr$gene, ], 
         name = "pre-mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = FALSE,
         column_title = "Intronic Expression",
        # column_order = cols, 
         split = time_to_expr$stage_expression, 
        row_title_rot = 0)
ht_pre

ht_gene <- Heatmap(sig_gene_log_counts[time_to_expr$gene, ], 
         name = "mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Exonic Expression",
         cluster_row_slices = FALSE,
        # column_order = cols, 
         split = time_to_expr$stage_expression, 
         row_title_rot = 0)
ht_gene

ht_pre + ht_gene

pdf(file.path(fig_dir, "premrnas_heatmap_logexpr_time_exonic_staged.pdf"), 
    height = 7, 
    width = 7)
  draw(ht_gene + ht_pre)
dev.off()
```

Show average profiles

```{r}
pre_expr <- sig_pre_gene_log_counts %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
    t() %>% 
  scale() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
  geom_point(aes(color = count_type)) +
  geom_line(aes(color = count_type, group = count_type)) + 
  facet_wrap(~stage_expression)
```


### Based on intronic read increases 

```{r order_by_time_expr_intronic}

gene_class_selected <- filter(gene_class_df, 
                              mzt_class %in% c(
                                 "zygotic_genes_intron"
                              ))

sig_pre_gene_counts <- mean_tpms_rissland %>% 
  filter(str_detect(gene_id, "-I[0-9]*$"),
         str_remove(gene_id, "-I[0-9]*$") %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = str_remove(gene_id, "-I[0-9]*$")) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix()

sig_pre_gene_log_counts <- sig_pre_gene_counts %>% 
  log1p()

sig_gene_log_counts <- mean_tpms_rissland %>% 
  filter(!str_detect(gene_id, "-I[0-9]*$"),
         gene_id %in% gene_class_selected$gene_id) %>% 
 # mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() %>% 
  log1p()

time_to_expr <- sig_pre_gene_counts %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>% 
  gather(stage, expr, -gene) %>% 
  as_tibble() %>% 
  mutate(expressed = expr > 0.1) %>% 
  mutate(stage = factor(stage,
                        levels = c(
                                           "0-1 hr embryo",
                                           "1-2 hr embryo",
                                           "2-3 hr embryo",
                                           "3-4 hr embryo",
                                           "4-5 hr embryo")
                                         )) %>% 
  group_by(gene) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  ungroup() %>% 
  filter(expressed) %>% 
  group_by(gene) %>% 
  mutate(stage_expression = dplyr::first(stage))
            
time_to_expr <- left_join(data_frame(gene = rownames(sig_pre_gene_log_counts)),
                          time_to_expr, by = "gene") %>% 
  na.omit()

ht_pre <- Heatmap(sig_pre_gene_log_counts[time_to_expr$gene, ], 
         name = "pre-mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = FALSE,
         column_title = "Intronic Expression",
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_pre

ht_gene <- Heatmap(sig_gene_log_counts[time_to_expr$gene, ], 
         name = "mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Exonic Expression",
         cluster_row_slices = FALSE,
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_gene

pdf(file.path(fig_dir, "premrnas_heatmap_logexpr_time_intronic_staged.pdf"), 
    height = 7, 
    width = 7)
  draw(ht_gene + ht_pre)
dev.off()
```

Show average profiles

```{r}
pre_expr <- sig_pre_gene_log_counts %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
    t() %>% 
  scale() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
  geom_point(aes(color = count_type)) +
  geom_line(aes(color = count_type, group = count_type)) + 
  facet_wrap(~stage_expression)
```

### Only zygotic (pure zygotic)

```{r order_by_time_expr_purezyg}

gene_class_selected <- filter(gene_class_df, 
                              mzt_class %in% c(
                                 "pure_zygotic_combined"
                              ))
sig_pre_gene_counts <- mean_tpms_rissland %>% 
  filter(str_detect(gene_id, "^pre"),
         no_pre(gene_id) %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() 
  
sig_pre_gene_log_counts <- sig_pre_gene_counts %>% 
  log1p()

sig_gene_log_counts <- mean_tpms_rissland %>% 
  filter(!str_detect(gene_id, "^pre"),
         no_pre(gene_id) %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() %>% 
  log1p()


time_to_expr <- sig_pre_gene_counts %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>% 
  gather(stage, expr, -gene) %>% 
  as_tibble() %>% 
  mutate(expressed = expr > 0.1) %>% 
  mutate(stage = factor(stage,
                        levels = c(
                                           "0-1 hr embryo",
                                           "1-2 hr embryo",
                                           "2-3 hr embryo",
                                           "3-4 hr embryo",
                                           "4-5 hr embryo")
                                         )) %>% 
  group_by(gene) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  ungroup() %>% 
  filter(expressed) %>% 
  group_by(gene) %>% 
  summarize(stage_expression = dplyr::first(stage))
            
time_to_expr <- left_join(data_frame(gene = rownames(sig_pre_gene_log_counts)),
                          time_to_expr, by = "gene") %>% 
  na.omit()

# sig_pre_gene_log_counts <- t(scale(t(sig_pre_gene_log_counts)))
# sig_gene_log_counts <- t(scale(t(sig_gene_log_counts)))
# 
# exclude_missing <- rownames(sig_gene_log_counts)[rowSums(is.na(sig_gene_log_counts[time_to_expr$gene, ])) == 0]
# 
# time_to_expr <- time_to_expr %>% filter(gene %in% exclude_missing)
# 
# sig_pre_gene_log_counts <- sig_pre_gene_log_counts[time_to_expr$gene, ]   
# sig_gene_log_counts <- sig_gene_log_counts[time_to_expr$gene, ]   

ht_pre <- Heatmap(sig_pre_gene_log_counts[time_to_expr$gene, ], 
         name = "pre-mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Intronic Expression",
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_pre

ht_gene <- Heatmap( sig_gene_log_counts[time_to_expr$gene, ], 
         name = "mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Exonic Expression",
         cluster_row_slices = FALSE,
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_gene

pdf(file.path(fig_dir, "premrnas_heatmap_logexpr_time_purezygotic_staged.pdf"), 
    height = 7, 
    width = 7)
  draw(ht_gene + ht_pre)
dev.off()
```


Show average profiles

```{r}
pre_expr <- sig_pre_gene_log_counts %>% 
  # t() %>% 
  # scale() %>% 
  # t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
  #   t() %>% 
  # scale() %>% 
  # t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

pavg_expr <- expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
    geom_point(aes(color = count_type)) +
    geom_line(aes(color = count_type, group = count_type)) + 
    scale_color_viridis_d(name = "") +
    facet_wrap(~stage_expression, ncol = 1) +
    labs(x = "", y = "Average Expression")  +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5))

pre_expr <- sig_pre_gene_log_counts %>% 
  t() %>%
  scale() %>%
  t() %>%
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
    t() %>%
  scale() %>%
  t() %>%
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

pavg_zscore <- expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
    geom_point(aes(color = count_type)) +
    geom_line(aes(color = count_type, group = count_type)) + 
    scale_color_viridis_d(name = "") +
    facet_wrap(~stage_expression, ncol = 1) +
    labs(x = "", y = "Average Expression (z-score)") +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5))

p <- plot_grid(pavg_expr, pavg_zscore, nrow = 1, ncol = 2)
p
save_plot(file.path(fig_dir, "avergage_expr_time_purezygotic_staged.pdf"),
          p,
          base_height = 7.5,
          base_asp = 0.66,
          nrow = 1,
          ncol = 2)
```

### All zygotic (all zygotic)

```{r order_by_time_expr_allzyg}

gene_class_selected <- filter(gene_class_df, 
                              mzt_class %in% c(
                                 "all_zygotic_combined"
                              ))
sig_pre_gene_log_counts <- mean_tpms_rissland %>% 
  filter(str_detect(gene_id, "^pre"),
         no_pre(gene_id) %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() %>% 
  log1p() 

sig_gene_log_counts <- mean_tpms_rissland %>% 
  filter(!str_detect(gene_id, "^pre"),
         no_pre(gene_id) %in% gene_class_selected$gene_id) %>% 
  mutate(gene_id = no_pre(gene_id)) %>% 
  tibble::column_to_rownames("gene_id") %>% 
  as.matrix() %>% 
  log1p() 


time_to_expr <- sig_pre_gene_log_counts %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>% 
  gather(stage, expr, -gene) %>% 
  as_tibble() %>% 
  mutate(expressed = expr > 0.5) %>% 
  mutate(stage = factor(stage,
                        levels = c(
                                           "0-1 hr embryo",
                                           "1-2 hr embryo",
                                           "2-3 hr embryo",
                                           "3-4 hr embryo",
                                           "4-5 hr embryo")
                                         )) %>% 
  group_by(gene) %>% 
  arrange(stage, .by_group = TRUE) %>% 
  ungroup() %>% 
  filter(expressed) %>% 
  group_by(gene) %>% 
  summarize(stage_expression = dplyr::first(stage))
            
time_to_expr <- left_join(data_frame(gene = rownames(sig_pre_gene_log_counts)),
                          time_to_expr, by = "gene") %>% 
  na.omit()

ht_pre <- Heatmap(sig_pre_gene_log_counts[time_to_expr$gene, ], 
         name = "pre-mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Intronic Expression",
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_pre

ht_gene <- Heatmap(sig_gene_log_counts[time_to_expr$gene, ], 
         name = "mRNA\nlog expression",
         show_row_names = FALSE,
         show_row_dend = FALSE,
         cluster_columns = FALSE,
         cluster_rows = TRUE,
         column_title = "Exonic Expression",
         cluster_row_slices = FALSE,
        # column_order = cols, 
         split = time_to_expr$stage_expression, row_title_rot = 0)
ht_gene

pdf(file.path(fig_dir, "premrnas_heatmap_logexpr_time_allzygotic_staged.pdf"), 
    height = 7, 
    width = 7)
  draw(ht_gene + ht_pre)
dev.off()
```

Show average profiles

```{r}
pre_expr <- sig_pre_gene_log_counts %>% 
  # t() %>% 
  # scale() %>% 
  # t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
  #   t() %>% 
  # scale() %>% 
  # t() %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

pavg_expr <- expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
    geom_point(aes(color = count_type)) +
    geom_line(aes(color = count_type, group = count_type)) + 
    scale_color_viridis_d(name = "") +
    facet_wrap(~stage_expression, ncol = 1) +
    labs(x = "", y = "Average Expression")  +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5))

pre_expr <- sig_pre_gene_log_counts %>% 
  t() %>%
  scale() %>%
  t() %>%
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "pre-mRNA")

mat_expr <- sig_gene_log_counts %>% 
    t() %>%
  scale() %>%
  t() %>%
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  inner_join(time_to_expr, by = "gene") %>% 
  mutate(count_type = "mRNA")

expr_vals <- bind_rows(pre_expr, mat_expr)

pavg_zscore <- expr_vals %>% 
  pivot_longer(cols = -c(gene, count_type, stage_expression)) %>% 
  group_by(stage_expression, count_type, name) %>%
  summarize(avg_expr = mean(value, na.rm = TRUE)) %>% 
  ggplot(aes(name, avg_expr)) +
    geom_point(aes(color = count_type)) +
    geom_line(aes(color = count_type, group = count_type)) + 
    scale_color_viridis_d(name = "") +
    facet_wrap(~stage_expression, ncol = 1) +
    labs(x = "", y = "Average Expression (z-score)") +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5))

p <- plot_grid(pavg_expr, pavg_zscore, nrow = 1, ncol = 2)
p
# save_plot(file.path(fig_dir, "avergage_expr_time_purezygotic_staged.pdf"),
#           p,
#           base_height = 7.5,
#           base_asp = 0.66,
#           nrow = 1,
#           ncol = 2)
```

```{r}
fn <- "../../docs/bartel_2019_s1.xlsx"

stages <- c(
  "NC_7-9",
  "NC_9-10",
  "SYNCYTIAL_BLASTODERM",
  "CELLULARIZED_BLASTODERM")

b_dat <- read_excel(fn, skip = 2)

b_genes <- b_dat %>%
  dplyr::select(GENES, one_of(stages)) %>% 
  left_join(gene2symbol, by = c("GENES" = "gene_name")) %>% 
  na.omit() 

# get stage specific expression
b_gene_stages <- b_genes %>% 
  pivot_longer(cols = -c(GENES, gene_id)) %>%
  filter(value) %>% 
  mutate(stages = factor(name, levels = stages)) %>% 
  group_by(gene_id) %>% 
  arrange(stages, .by_group = TRUE) %>% 
  summarize(tpt = dplyr::first(stages)) %>% 
  split(., .$tpt)
```


## Compare to bartel data per timepoint

```{r}
time_to_expr$stage_expression <- droplevels(time_to_expr$stage_expression)
# for(i in seq_along(levels(time_to_expr$stage_expression))){
#   tpt <- levels(time_to_expr$stage_expression)[i]
#   map(stages, ~{
#     to_comp <- list(
#     `Kwasnieski et al` = b_gene_stages[[.x]]$gene_id,
#     `Intron method` = time_to_expr %>% filter(stage_expression == tpt) %>% pull(gene)
#   ) %>% 
#     map(~na.omit(.x) %>% .[!duplicated(.)]) 
#   
# a <- euler(to_comp)
# 
# p <- plot(a, main = tpt,
#           quantities = list(fontsize = 16), 
#           labels = list(fontsize = 20))
# print(p)
#   })} 

plt_order <- c(
  "NC_7-9",
  "NC_9-10",
  "SYNCYTIAL_BLASTODERM",
  "CELLULARIZED_BLASTODERM",
  "Not Observed")

plt_dat <- left_join(time_to_expr,
          b_gene_stages %>% bind_rows(), 
          by = c("gene" = "gene_id")) %>% 
  mutate(timepoint = ifelse(is.na(tpt), "Not Observed", as.character(tpt)),
         timepoint = factor(timepoint, levels = plt_order)) %>% 
  group_by(stage_expression) %>% 
  mutate(n_genes = n()) %>% 
  group_by(stage_expression, timepoint) %>% 
    mutate(timepoint_prop = n() / n_genes) %>% 
  ungroup()
  
p1 <- ggplot(plt_dat, aes(stage_expression)) + 
  geom_bar(aes(fill = timepoint)) +
  scale_fill_brewer(palette = "Set1", name = "Kwasnieski et al.\n classes") +
  labs(x = "", y = "# of genes") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "none")


plt_dat_prop<- plt_dat %>% 
  select(stage_expression, timepoint, timepoint_prop) %>%
  unique() 

plt_dat_text <- plt_dat %>% 
  select(stage_expression, n_genes) %>%
  mutate(n_genes = str_c("n = ", n_genes)) %>% 
  unique() 


p2 <- ggplot(plt_dat_prop, aes(stage_expression, timepoint_prop)) + 
  geom_col(aes(fill = timepoint)) +
  geom_text(data = plt_dat_text, aes(y = 0.2, label = n_genes), angle = 90) +
  scale_fill_brewer(palette = "Set1", name = "Kwasnieski et al.\nclasses") +
  labs(x = "", y = "Proportion of Genes") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p <- plot_grid(p1, p2, nrow = 1, ncol = 2, rel_widths = c(0.5, 0.5))

p

save_plot(file.path(fig_dir, "kwasnieksi_comp_counts.pdf"), 
          p1, base_height = 3.25, base_width = 2.5)

save_plot(file.path(fig_dir, "kwasnieksi_comp_prop.pdf"), 
          p2, base_asp = 1.4)
```


```{r}
out <- dplyr::rename(time_to_expr, gene_id = gene) %>% 
  left_join(.,gene2symbol, by = "gene_id") %>% 
  arrange(stage_expression) %>% 
  left_join(filter(out_tbl, count_type == "intron"))

write_tsv(out, file.path(outdir, "time_to_expression.tsv"))


gtf <- tidy_gtf(annots$drosophila$gtf)

```

### # of introns per class


```{r}
intron_dat <- full_join(time_to_expr,
          b_gene_stages %>% bind_rows(), 
          by = c("gene" = "gene_id")) %>% 
  mutate(timepoint = ifelse(is.na(tpt), "Not Observed", as.character(tpt)),
         stage_expression = ifelse(is.na(stage_expression),
                                   "Not Observed", 
                                   as.character(stage_expression)),
         timepoint = factor(timepoint, levels = plt_order)) %>% 
  group_by(stage_expression) %>% 
  mutate(n_genes = n()) %>% 
  group_by(stage_expression, timepoint) %>% 
    mutate(timepoint_prop = n() / n_genes) %>% 
  ungroup() %>% 
  dplyr::rename(gene_id = gene) %>% 
  left_join(.,gene2symbol, by = "gene_id") %>% 
  arrange(stage_expression) 



n_introns_per_gene <- gtf %>% 
  filter(type == "exon") %>% 
  group_by(transcript_id, gene_id) %>% 
  summarize(n_exons = n(), n_introns = n_exons - 1L) %>% 
  group_by(gene_id) %>% 
  summarize(n_introns = max(n_introns))


plt_dat <- intron_dat %>% 
  left_join(n_introns_per_gene, by = "gene_id") %>% 
  filter(timepoint != "Not Observed") %>% 
  mutate(seen_by_intron = ifelse(stage_expression == "Not Observed",
                                 "Not Observed",
                                 "Observed"))
  

p1 <- ggplot(plt_dat, aes(timepoint, n_introns)) + 
  geom_boxplot(aes(fill = seen_by_intron), coef = 1e9) +
  scale_fill_brewer(palette = "Set1", name = "Detected by\nIntronic reads") +
  labs(x = "", y = "# of introns") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


save_plot(file.path(fig_dir, "kwasnieksi_comp_introns.pdf"), 
          p1, base_asp = 0.75, base_height = 6)


plt_dat <- intron_dat %>% 
  left_join(n_introns_per_gene, by = "gene_id") %>% 
  filter(stage_expression != "Not Observed") %>% 
  mutate(seen_by_kwasnieksi = ifelse(timepoint == "Not Observed",
                                 "Not Observed",
                                 "Observed"))
  

p2 <- ggplot(plt_dat, aes(stage_expression, n_introns)) + 
  geom_boxplot(aes(fill = seen_by_kwasnieksi), coef = 1e9) +
  scale_fill_brewer(palette = "Set1", name = "Detected by\n4-SU") +
  labs(x = "", y = "# of introns") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p2
save_plot(file.path(fig_dir, "kwasnieksi_comp_introns_our_data.pdf"), 
          p2,base_height = 5, base_width = 5)

```


```{r eisen_zelda_dat}
# use combined data
zelda_chip <- read_xls("../../docs/eisen_zelda_chip_master_table.xlsx") %>% 
  dplyr::select(Gene, `ZLD Max cyc8`:`ZLD Max cyc14`) %>% 
  pivot_longer(cols = -Gene) %>% 
  group_by(Gene) %>% 
  summarize(zelda_signal = max(value)) %>% 
  left_join(gene2symbol, by = c("Gene" = "gene_name"))

plt_dat <- left_join(time_to_expr, zelda_chip, by = c("gene" = "gene_id")) %>% 
  na.omit()
  
plt_dat_no_zero <- filter(plt_dat, zelda_signal != 0)
plt_txt <- plt_dat_no_zero %>% group_by(stage_expression) %>% 
  summarize(n= n()) %>% 
  mutate(n = str_c("n = ", n))
  
p <- ggplot(plt_dat_no_zero, aes(stage_expression , zelda_signal)) + 
  geom_boxplot(aes(fill = stage_expression ),
                coef = Inf) +
  geom_text(data = plt_txt, aes(y = 11000, label = n)) +
  scale_y_log10() +
  labs(x = "",
       y = "Zelda Chip-Seq Signal") +
  scale_fill_viridis_d() +
  theme(legend.position = "none")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p
save_plot(file.path(fig_dir, "zelda-binding_by_tpt.pdf"),
          p,
          base_asp = 1.2)

plt_dat <- left_join(time_to_expr, zelda_chip, by = c("gene" = "gene_id")) %>% 
  na.omit() %>% 
  mutate(zelda_signal = zelda_signal + 1)

plt_txt <- plt_dat %>% group_by(stage_expression) %>% 
  summarize(n= n()) %>% 
  mutate(n = str_c("n = ", n))

p <- ggplot(plt_dat, aes(stage_expression , zelda_signal)) + 
  geom_violin(aes(fill = stage_expression)) +
  geom_text(data = plt_txt, aes(y = 11000, label = n)) +
  scale_y_log10() +
  labs(x = "",
       y = "Zelda Chip-Seq Signal") +
  scale_fill_viridis_d() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p
save_plot(file.path(fig_dir, "zelda-binding_no_rm_zero.pdf"),
          p,
          base_asp = 1.2)
```


